---
title: "Getting Started"
author: "<h4>Authors: <i>`r auths <- eval(parse(text = gsub('person','c',read.dcf('../DESCRIPTION', fields = 'Authors@R'))));paste(auths[names(auths)=='given'],auths[names(auths)=='family'], collapse = ', ')`</i></h4>"
date: "<h4>Vignette updated: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i></h4>"
output:
  BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{echoLD}
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`echoLD` is an [echoverse](https://github.com/RajLabMSSM/echoverse) module
for retrieving and processing linkage disequilibrium (LD) matrices.
It supports multiple LD reference panels:

- **1000 Genomes** Phase 1 and Phase 3 (computed on-the-fly from VCF)
- **UK Biobank** (pre-computed LD matrices)
- **Custom VCF** files supplied by the user
- **Pre-computed LD matrices** in `.rds`, `.csv`, or `.tsv` format

# Setup

```{r load}
library(echoLD)
```

The examples below use bundled data from the `echodata` package:
a subset of summary statistics from the *BST1* locus and a
pre-computed LD matrix.

```{r example_data}
query_dat <- echodata::BST1[seq_len(50), ]
locus_dir <- file.path(tempdir(), echodata::locus_dir)
LD_matrix <- echodata::BST1_LD_matrix
```

# Pre-computed LD matrix

If you already have an LD matrix on disk, `get_LD_matrix()` reads it in
and aligns it with your summary statistics.

```{r get_LD_matrix}
## Write the bundled LD matrix to a temp CSV to demonstrate the workflow
LD_path <- tempfile(fileext = ".csv")
utils::write.csv(LD_matrix, file = LD_path, row.names = TRUE)

LD_list <- echoLD::get_LD_matrix(
    locus_dir = locus_dir,
    query_dat = query_dat,
    LD_reference = LD_path
)
```

# Custom VCF

You can compute LD on-the-fly from any VCF file.
The `echodata` package ships a small 1000 Genomes Phase 3 VCF
for the BST1 locus.

```{r get_LD_vcf, eval = requireNamespace("VariantAnnotation", quietly = TRUE) && requireNamespace("snpStats", quietly = TRUE)}
LD_reference <- system.file("extdata", "BST1.1KGphase3.vcf.bgz",
                            package = "echodata")
LD_vcf <- echoLD::get_LD_vcf(
    locus_dir = locus_dir,
    query_dat = query_dat,
    LD_reference = LD_reference
)
```

# 1000 Genomes (remote)

Querying the full 1000 Genomes VCF requires downloading chromosome-level
VCF files from a remote server. This requires an internet connection
and can take several minutes.

```{r get_LD_1kg, eval = FALSE}
LD_1kgp3 <- echoLD::get_LD(
    locus_dir = locus_dir,
    query_dat = query_dat,
    LD_reference = "1KGphase3"
)
```

# UK Biobank (remote)

Pre-computed LD from a British European-descent subset of UK Biobank.
**Note:** This takes substantially longer than the 1000 Genomes methods.

```{r get_LD_ukb, eval = FALSE}
LD_ukb <- echoLD::get_LD(
    locus_dir = locus_dir,
    query_dat = query_dat,
    LD_reference = "UKB",
    download_method = "axel",
    nThread = 10
)
```

# Filtering LD

Use `filter_LD()` to remove SNPs below a minimum r2 threshold.

```{r filter_LD}
LD_list_full <- list(LD = LD_matrix, DT = echodata::BST1)
LD_list_filt <- echoLD::filter_LD(LD_list = LD_list_full, min_r2 = 0.2)
dim(LD_list_filt$LD)
```

# Sparse matrix utilities

`echoLD` provides helpers for working with sparse LD matrices,
which reduce file size substantially.

```{r sparse}
## Convert a dense matrix to sparse format
sparse_mat <- echoLD::to_sparse(X = LD_matrix)
class(sparse_mat)

## Save to disk
sparse_path <- echoLD::saveSparse(LD_matrix = LD_matrix)
file.info(sparse_path)$size

## Read it back
sparse_read <- echoLD::readSparse(LD_path = sparse_path)
dim(sparse_read)
```

# Plotting LD

Plot a heatmap of pairwise LD around the lead SNP.

```{r plot_LD, fig.width = 7, fig.height = 6}
echoLD::plot_LD(
    LD_matrix = LD_matrix,
    query_dat = echodata::BST1,
    span = 10
)
```

# Session Info

<details>

```{r Session Info}
utils::sessionInfo()
```

</details>

<br>
